Supporting Information:\\ Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-23 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
Full name, institution and contribution of FFR.
Tiny changes to the SI.

\documentclass[journal=jpcbfk]{achemso}

@@

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}

@@

\newpage
%\tableofcontents

@@

  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\

@@

Simulations of these systems were performed for 500 ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2 fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303 K and 298 K for DOPS and POPS respectively using the Nosé-Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1 ps. A pressure of 1 bar was maintained using the Parrinello-Rahman \cite{parrinello81} method with a coupling constant of 5 ps.
Pressure coupling was applied in a semi-isotropic manner with the x and y dimensions, in the plane of the bilayer, fluctuating independently of the z dimension.
Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2 nm with the interactions
switched off between 1.0 nm and 1.2 nm; Coulombic interactions were truncated at 1.2 nm with long-range interactions treating using PME~\cite{darden93,essman95}.

Analyses on these simulations was performed on the final 100 ns of the simulations.

@@

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
with neutralizing potassium countertions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
 \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
potassium or sodium ions corresponding concentrations of $\sim$450 mM and 890 mM
were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
parameters given by the CHARMM-GUI at 298 K. For further details, simulation files
and data, see table \ref{mixedIONsystemsMONOVALENT} and
Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} 
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26 M and 1 M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 298 K.

@@

\noindent
{\it POPS bilayers with Na$^+$ counterions.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the provided PS parameters as a guide~\cite{rog16}. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na+ ions) and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0 nm with a dispersion correction applied to the energy and pressure;
Coulombic interactions were truncated at 1.0 nm with PME used to treat interactions beyond this cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.

@@

The simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step. 
%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.

\subsection{Lipid17}
{\it DOPS and POPS bilayers.} 
The structure of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
4480 water molecules resulting in 31.1 water molecules per lipid. For
each system, 128 NaCl molecules are added to neutralize the system.
In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{amber18md}
and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
the number of molecules in simulations, related simulation files and trajectories.

Each bilayer structure was first minimized for 2500 steps with steepest
descent and an additional 2500 step with the conjugent gradient
algorithms while restrainig the solvent and ion molecules with a force
constant 500~kcal~mol$^{-1}$~\AA{}$^{-1}$. An additional minimization
step was used with the same parameters after removing the constraints on
the solvent and ion molecules. During the minimization procedure,
non-periodic boundary conditions were applied, no pressure scaling was
used and none of the bonds were restrained. A three step heating
procedure was applied: in the first heating step the system temperature
is increased from 0.0~K to 100.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency while restraining the lipid positions
with harmonic springs of force constant
20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume without periodic
boundary conditions. In the second heating step, the system temperature
was increased from 100.0~K to 200.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency within 10000 steps while restraining
the lipid positions with harmonic springs of force constant
10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume. In the third
heating step, the system temperature is increased from 200.0~K to
303.0~K within 10000 steps using Langevin thermostat with 5.0~ps$^{-1}$
collision frequency while restraining the lipid positions with harmonic
springs of force constant 10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at 1~bar
constant pressure using Berendsen barostat with 3.0~ps relaxation time.
From the second heating step onward, periodic boundary conditions in all
dimensions were applied. The semi-isotropic pressure coupling along
\textit{xy}-plane was employed and the surface tension was set to
0.0~dyn explicitely. After the third heating step, each structure is
equilibriated for 100~ns at 1~bar and 303~K constant pressure using
Berendsen barostat and temperature using Langevin thermostat,
respectively. For the equilibriation step, semi-isotropic pressure
coupling along \textit{xy}-plane was employed and the surface tension
was set to 0.0~dyn explicitely. For all heating and equilibration steps,
the lengths of the hydrogen containing bonds were restrained using SHAKE
algorithm and a time step of 2~fs was used. In all steps with periodic
boundary conditions, the non-bonded interactions are calculated with
particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with Dang ion parameters} 
The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
and converted to Gromacs format using acpype \cite{silva12}.  
The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
Dang parameters for ions \cite{smith94, dang06} were used.
% Another set of simulations with Lipid17 force field for lipids \cite{gould18}
Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18} 
and simulations were performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}
with the time step of 2~fs. 
The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
The bond lengths of hydrogen atoms were constrained using LINCS algorithm \cite{hess97}.
Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
Pressure was coupled to the Parrinello-Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
The systems were equilibrated 50~ns prior to production simulations used for analysis,
which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
Topologies in a format for the GROMACS package are available in the repository \citenum{ECCpops}.

@@

\noindent
{\it DOPS with 288 lipids.}
The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na$^+$ ions
was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
The Slipids~\cite{jambeck13} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
ions are described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
The system was simulated in NPT ensemble for 200~ns using the GROMACS 5.0.4 package~\cite{abraham2015gromacs},
and the last 100~ns was used for the analysis.
Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

\subsection{Berger}
\noindent
{\it POPC bilayers.} 
Previously published simulation \cite{ollila07a} available from Ref. \citenum{bergerPOPC310K}
was used for POPC at 310 K.

\noindent
{\it DOPS and POPS bilayers.} 
%The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
%\todo{T. Piggot, please write simulation details} \\
Starting structures for DOPS and POPS simulations performed with a Berger-based PS force field
were taken from those produced for the CHARMM36 simulations. A script was used to re-order the
atoms and to remove any extraneous hydrogen atoms. Force field parameters were taken from those
published by Mukhopadhyay et al.~\cite{mukhopadhyay04}
% [REF: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1303994/]
and the SPC water model was used~\cite{berendsen81}. The simulation cut-offs closely matched
those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0 nm;
Coulombic interactions were truncated at 1.0 nm with PME applied to account for the interactions beyond the cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.

@@

That said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
bound charge are typically smaller than then discrepancies arising from ion binding affinity.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer from simulations and experiments \cite{scherer89}.
  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
}
\end{figure}

@@

\noindent
{\it Fernando Favela}

>>>>

\documentclass[journal=jctcce]{achemso}

@@

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela-Rosales}
\affiliation{Departamento de Investigaci\'{o}n, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}

@@

%\newpage
%\tableofcontents

@@

  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\

@@

Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303~K and 298~K for DOPS and POPS respectively using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}.

Analyses on these simulations was performed on the final 100~ns of the simulations.

@@

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
with neutralizing potassium countertions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
 \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
potassium or sodium ions corresponding concentrations of $\sim$450~mM and 890~mM
were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
and data, see table \ref{mixedIONsystemsMONOVALENT} and
Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} 
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 298~K.

@@

\noindent
{\it POPS bilayers with Na$^+$ counterions.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the provided PS parameters as a guide~\cite{rog16}. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions) and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
Long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.

@@

The simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step. 
%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for long-range electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.

\subsection{Lipid17}
{\it DOPS and POPS bilayers.} 
The structure of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
4480 water molecules resulting in 31.1 water molecules per lipid. For
each system, 128 NaCl molecules are added to neutralize the system.
In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{amber18md}
and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung--Cheatham
ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
the number of molecules in simulations, related simulation files and trajectories.

Each bilayer structure was first minimized for 2500 steps with steepest
descent and an additional 2500 step with the conjugate gradient
algorithms while restraining the solvent and ion molecules with a force
constant 500~kcal~mol$^{-1}$~\AA{}$^{-2}$. An additional minimization
step was used with the same parameters after removing the constraints on
the solvent and ion molecules. During the minimization procedure,
non-periodic boundary conditions were applied, no pressure scaling was
used and none of the bonds were restrained. A three step heating
procedure was applied: in the first heating step the system temperature
is increased from 0.0~K to 100.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency while restraining the lipid positions
with harmonic springs of force constant
20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume without periodic
boundary conditions. In the second heating step, the system temperature
was increased from 100.0~K to 200.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency within 10000 steps while restraining
the lipid positions with harmonic springs of force constant
10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume. In the third
heating step, the system temperature is increased from 200.0~K to
303.0~K within 10000 steps using Langevin thermostat with 5.0~ps$^{-1}$
collision frequency while restraining the lipid positions with harmonic
springs of force constant 10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at 1~bar
constant pressure using Berendsen barostat with 3.0~ps relaxation time.
From the second heating step onward, periodic boundary conditions in all
dimensions were applied. The semi-isotropic pressure coupling along
\textit{xy}-plane was employed and the surface tension was set to
0.0~dyn explicitely. After the third heating step, each structure is
equilibriated for 100~ns at 1~bar and 303~K constant pressure using
Berendsen barostat and temperature using Langevin thermostat,
respectively. For the equilibriation step, semi-isotropic pressure
coupling along \textit{xy}-plane was employed and the surface tension
was set to 0.0~dyn explicitely. For all heating and equilibration steps,
the lengths of the hydrogen containing bonds were restrained using SHAKE
algorithm and a time step of 2~fs was used. In all steps with periodic
boundary conditions, the non-bonded interactions are calculated with
particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with Dang ion parameters} 
The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
and converted to Gromacs format using acpype \cite{silva12}.  
The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
Dang parameters for ions \cite{smith94, dang06} were used.
% Another set of simulations with Lipid17 force field for lipids \cite{gould18}
Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18} 
and simulations were performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}
with the time step of 2~fs. 
The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
The bond lengths of hydrogen atoms were constrained using LINCS algorithm \cite{hess97}.
Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
Pressure was coupled to the Parrinello--Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
The systems were equilibrated 50~ns prior to production simulations used for analysis,
which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
Topologies in a format for the GROMACS package are available in the repository \citenum{ECCpops}.

@@

\noindent
{\it DOPS with 288 lipids.}
The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na$^+$ ions
was constructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
The Slipids~\cite{jambeck13} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
ions are described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
The system was simulated in NPT ensemble for 200~ns using the GROMACS 5.0.4 package~\cite{abraham2015gromacs},
and the last 100~ns was used for the analysis.
Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi-isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

\subsection{Berger}
\noindent
{\it POPC bilayers.} 
Previously published simulation \cite{ollila07a} available from Ref. \citenum{bergerPOPC310K}
was used for POPC at 310~K.

\noindent
{\it DOPS and POPS bilayers.} 
%The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
%\todo{T. Piggot, please write simulation details} \\
Starting structures for DOPS and POPS simulations performed with a Berger-based PS force field
were taken from those produced for the CHARMM36 simulations. A script was used to re-order the
atoms and to remove any extraneous hydrogen atoms. Force field parameters were taken from those
published by Mukhopadhyay et al.~\cite{mukhopadhyay04}
% [REF: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1303994/]
and the SPC water model was used~\cite{berendsen81}. The simulation cut-offs closely matched
those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0~nm;
Coulombic interactions were truncated at 1.0~nm with PME applied to account for the interactions beyond the cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.

@@

That being said, in the force fields investigated so far, the discrepancies arising from the sentivity of lipid headgroup to
bound charge are typically smaller than then discrepancies arising from ion binding affinity.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer from simulations and experiments \cite{scherer89}.
  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
}
\end{figure}

@@

\noindent
{\it Fernando Favela-Rosales}
Setup and performed one of the DOPS simulations with Slipids.
